---
title: "hw01_gapminder.Rmd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Seevasant Indran
output: github_document
keep_md: TRUE
---

Lets explore the gapminder dataset.

## Getting Started with Gapminder

```{r echo=FALSE, eval=FALSE}
#Install the gapminder package
install.packages("gapminder")
```

```{r, message=FALSE}
#Load the gapminder dataset and the tidyverse library.
library(gapminder)
library(tidyverse)

```

## Lets have a look at the Gapminder dataset


```{r, eval=F}
# Lets see what is the gapminder data is all about in a separate window
View(gapminder) 

```

```{r}

#lets assign the gapminder dataset to the variable hw01gapminder, number of onservation (rows = 1704) and variables (column = 6), look on the global environment.

hw01gapminder <- gapminder

# check if the assigned data is true in the original dataset using the ncol() and nrow() function.

ncol(gapminder) == ncol(hw01gapminder)
ncol(gapminder) == ncol(hw01gapminder)
```

```{r}

#View the top and the botttom 6 of the gapminder dataset
  
head(hw01gapminder)
tail(hw01gapminder)

```

## Lets plot the data

```{r}
#Using ggplot to plot life expectancy vs year
ggplot(data = hw01gapminder) + 
  geom_point(mapping = aes(x = year, y = lifeExp))
```

It seems that the life epectancy has increased over the last 50 years.

```{r}
#Using ggplot to plot life expectancy vs year using colour to separate by continents
ggplot(data = hw01gapminder) + 
  geom_point(mapping = aes(x = year, y = lifeExp, color = continent))
```

The African continent has the lowest life expectancy and Europe has the highest life expentancy.

```{r}
ggplot(data = hw01gapminder) + 
  geom_point(mapping = aes(x = year, y = lifeExp, color = continent, size = pop))
```

Asia has the highest population although it can be seen the population in Asia has gotten smaller in recent years.

```{r}
ggplot(data = hw01gapminder) + 
  geom_point(mapping = aes(x = year, y = lifeExp, color = year))
```


