---
title: "hw01_gapminder"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Seevasant Indran
output: github_document
keep_md: TRUE
---

Lets explore the gapminder dataset of hw01 part 2

## Getting Started with Gapminder

```{r echo=FALSE, eval= FALSE}
#Install the gapminder package, set eval = FALSE and echo = FALSE so it doesnt run the code everytime the script is knit. 

install.packages("gapminder")
install.packages("tidyverse")

```

```{r, message=FALSE}
#Load the gapminder dataset and the tidyverse library.

library(gapminder)
library(tidyverse)

```

## Lets explore the Gapminder dataset structure

```{r, eval=F}
# First lets see what is the gapminder data is all about in a separate window

View(gapminder) 

```

```{r}

# lets assign the gapminder dataset to the variable hw01gapminder.

hw01gapminder <- gapminder # Number of observation (rows = 1704) and variables (column = 6), look on the global environment

# Checking class type
class (hw01gapminder) 


# Help about getting it to show more rows, looks like it the default displays 6 rows
?head() 

# lets set n = 10, so it will display 10 rows from the top
head(hw01gapminder, n = 10) 

# Show 6 rows from the bottom
tail(hw01gapminder)

# Is "country" a character ?
is.character(hw01gapminder$country)

# Is "country"" a factor ? 
is.factor(hw01gapminder$country) 

# Structure of the gapminder data
str(hw01gapminder)

# Dimensions of the dataframe (nrows x ncols)
dim(hw01gapminder) 

# Let see if the dim() output matches the ncol() and nrow() function.

nrow(hw01gapminder)
ncol(hw01gapminder)

# Getting "fancy" with checking if dimensions match row/column number.

(dim(hw01gapminder)[1]) == nrow(hw01gapminder)
(dim(hw01gapminder)[2]) == ncol(hw01gapminder)


# What variables are measured ?
colnames(hw01gapminder)

# What are the observations (countries) ?
levels(hw01gapminder$country)

# Limit for summary is 100 factors lets change it the number of factors. Each country has data for 12 years
summary(hw01gapminder$country, maxsum = 142)

# Is Malaysia in the list ? 
"Malaysia" %in% hw01gapminder$country


```

## More data exploration

```{r}
# Highest value for year
max(hw01gapminder$year) 

# Minimum value for year
min(hw01gapminder$year) 
```

```{r}

# Maximum value of age
max(hw01gapminder$lifeExp) 

# Minimum value of age
min(hw01gapminder$lifeExp) 

```

```{r}

# Maximum value of GDP per capita
max(hw01gapminder$gdpPercap) 

# Minimum value of GDP per capita
min(hw01gapminder$gdpPercap) 

# An overview of the statistics
summary(hw01gapminder)

```


## Malaysia gapminder data

```{r}

# I am from Malaysia, lets see how is Malaysia doing
GapminderMalaysia <- hw01gapminder %>% filter(country == "Malaysia")
head(GapminderMalaysia, n= 12)
```

```{r}

# Lets figure out which country has the highest life expectancy 
arrange(gapminder, desc(lifeExp))

# Lets pull out Malaysia and Japan to compare them
GapminderJapMas <- hw01gapminder %>% filter(country == "Malaysia" | country == "Japan")

# Lets look at the stats
summary(GapminderJapMas)

# The levels from the old dataframe are still retained, lets drop them out and try summary() again
GapminderJapMas<-droplevels(GapminderJapMas)
summary(GapminderJapMas)
```

It is not very clear from the tables, let's visualize these data in the next section.

## Plotting the data for visualization

```{r}

#Using ggplot to plot life expectancy vs year for Malaysians
ggplot(data = GapminderMalaysia) + 
  geom_point(mapping = aes(x = year, y = lifeExp))

```
  

I guess I will live another day to do more stats 540. 

### Malaysia vs Japan life expectancy

```{r}

# Box plot displays the data above better
ggplot(data = GapminderJapMas) + 
  geom_boxplot(mapping = aes(x = pop, y = lifeExp, color = country)) + coord_flip()

```

Not even a close fight, lets explore in terms of GDP and life expectancy throughout the years.


```{r}
ggplot(data = GapminderJapMas) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = country))
```

Japan has a very linear GDP growth. There seems to be some correlation of GDP per capita on age, lets try to see if this claim holds across all the continent. 


```{r}
ggplot(data = hw01gapminder) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = continent))
```

```{r}
arrange(hw01gapminder, desc(gdpPercap, lifeExp))
```


They seem to hold, looks like GDP per capita has an effect on health except for Kuwait. [Kuwait life expectancy article](https://www.ndtv.com/world-news/average-life-expectancy-of-indians-in-kuwait-way-below-national-figures-684817)


### World GDP per capita vs Life Expectancy


```{r}
# The outliers (Kuwait) is throwing the plots off, making it hard to viusalize the data, lets try dropping it of out data to see if it has a more even visualization

GapminderNoKuwait <- hw01gapminder %>% filter(!country=="Kuwait")
droplevels(GapminderNoKuwait)
```

### World GDP per capita vs Life Expectancy (without Kuwait)

```{r}
ggplot(data = GapminderNoKuwait) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = continent))
```

There seem to be some correlation in GDP per capita vs Life Expectancy


### World life expectancy vs Population

```{r}

# Boxplot life expectancy in different continents
ggplot(data = hw01gapminder) + 
  geom_boxplot(mapping = aes(x = continent, y = lifeExp)) + coord_flip()

```
The African continent has the lowest median life expectancy and Australia and Europe has the highest life expentancy and the smallest population. 


```{r}

# Boxplot life expectancy vs population in different continents

ggplot(data = hw01gapminder) + 
  geom_boxplot(mapping = aes(x = pop, y = lifeExp, color = continent))  + coord_flip()
```

Australia has the smallest range of life expectancy from 1952 - 2007 and the smallest population, Asia has the largest range of life Expectancy of population.
