---
title: "hw01_gapminder"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Seevasant Indran
output: github_document
keep_md: TRUE
---

Lets explore the gapminder dataset of hw01 part 2

## Getting Started with Gapminder

```{r echo=FALSE, eval= FALSE}
#Install the gapminder package, set eval = FALSE and echo = FALSE so it doesnt run the code everytime the script is knit. 

install.packages("gapminder")
install.packages("tidyverse")

```

```{r, message=FALSE}
#Load the gapminder dataset and the tidyverse library.

library(gapminder)
library(tidyverse)

```

## Lets explore the Gapminder dataset structure

```{r, eval=F}
# First lets see what is the gapminder data is all about in a separate window

View(gapminder) 

```

```{r}

# lets assign the gapminder dataset to the variable hw01gapminder.

hw01gapminder <- gapminder # Number of observation (rows = 1704) and variables (column = 6), look on the global environment

# Checking class type
class (hw01gapminder) 


# Help about getting it to show more rows, looks like it the default displays 6 rows
?head() 

# lets set n = 10, so it will display 10 rows from the top
head(hw01gapminder, n = 10) 

# Show 6 rows from the bottom
tail(hw01gapminder)

# Is "country" a character ?
is.character(hw01gapminder$country)

# Is "country"" a factor ? 
is.factor(hw01gapminder$country) 

# Structure of the gapminder data
str(hw01gapminder)

# Dimensions of the dataframe (nrows x ncols)
dim(hw01gapminder) 

# Let see if the dim() output matches the ncol() and nrow() function.

nrow(hw01gapminder)
ncol(hw01gapminder)

# Getting "fancy" with checking if dimensions match row/column number.

(dim(hw01gapminder)[1]) == nrow(hw01gapminder)
(dim(hw01gapminder)[2]) == ncol(hw01gapminder)


# What variables are measured ?
colnames(hw01gapminder)

# What are the observations (countries) ?
levels(hw01gapminder$country)

# Limit for summary is 100 factors lets change it the number of factors. Each country has data for 12 years
summary(hw01gapminder$country, maxsum = 142)

# Is Malaysia in the list ? 
"Malaysia" %in% hw01gapminder$country


```

## More data exploration

```{r}
# Highest value for year
max(hw01gapminder$year) 

# Minimum value for year
min(hw01gapminder$year) 
```

```{r}

# Maximum value of age
max(hw01gapminder$lifeExp) 

# Minimum value of age
min(hw01gapminder$lifeExp) 

```

```{r}

# Maximum value of GDP per capita
max(hw01gapminder$gdpPercap) 

# Minimum value of GDP per capita
min(hw01gapminder$gdpPercap) 

# An overview of the statistics
summary(hw01gapminder)

```


## Malaysia gapminder data

```{r}

# I am from Malaysia, lets see how is Malaysia doing
GapminderMalaysia <- hw01gapminder %>% filter(country == "Malaysia")
head(GapminderMalaysia, n= 12)
```

```{r}

# Lets figure out which country has the highest life expectancy 
arrange(gapminder, desc(lifeExp))

# Lets pull out Malaysia and Japan to compare them
GapminderJapMas <- hw01gapminder %>% filter(country == "Malaysia" | country == "Japan")

# Lets look at the statistics
summary(GapminderJapMas)

# The levels from the old dataframe are still retained, lets drop them out and try summary() again
GapminderJapMas<-droplevels(GapminderJapMas)
summary(GapminderJapMas)
```

It is not very clear from these, let's visualize these data in the next section.

## Plotting the data for visualization

```{r}

#Using ggplot to plot life expectancy vs year for Malaysians
ggplot(data = GapminderMalaysia) + 
  geom_point(mapping = aes(x = year, y = lifeExp))

```
  
I guess I will live another day to do more stats 540. 

### Malaysia vs Japan life expectancy

```{r}

# Box plot displays the data above better
ggplot(data = GapminderJapMas) + 
  geom_boxplot(mapping = aes(x = pop, y = lifeExp, color = country)) + coord_flip()

```

The median for Japanese is ~76 years old and the median for Malaysians is ~66 years old. 10 years !!, those sushi's must have something in them. Next, lets explore in terms of GDP and life expectancy throughout the years.


```{r}
ggplot(data = GapminderJapMas) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = country))
```

Japan has a very linear GDP growth. There seems to be some form of correlation between GDP per capita and life expectancy, lets try to see if this holds across all the continent. 


```{r}
ggplot(data = hw01gapminder) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = continent))
```

```{r}
# Finding the asian outliers on the plot
arrange(hw01gapminder, desc(gdpPercap, lifeExp))
```


Looks like GDP per capita has an effect on health except in Kuwait, but it is hard to say with the data squeezed onto the left side of the plot. [Kuwait life expectancy article](https://www.ndtv.com/world-news/average-life-expectancy-of-indians-in-kuwait-way-below-national-figures-684817)


### World GDP per capita vs Life Expectancy


```{r}
# The outliers (Kuwait) is making it hard to viusalize the data, lets try dropping it of out the dataframe to see if it imporves the plot visualization using the (!= not "Kuwait" operator)

GapminderNoKuwait <- hw01gapminder %>% filter(!country=="Kuwait")
droplevels(GapminderNoKuwait)
"Kuwait" %in% GapminderNoKuwait$country
```

### World GDP per capita vs Life Expectancy (without Kuwait)

```{r}
ggplot(data = GapminderNoKuwait) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color = continent))
```

There seem to be some correlation in GDP per capita vs Life Expectancy and it is clearer now that the plot is more spread out.

### Life expectancy (all continents)

```{r}

# Boxplot life expectancy in different continents
ggplot(data = hw01gapminder) + 
  geom_boxplot(mapping = aes(x = continent, y = lifeExp)) + coord_flip()

```
The African continent has the lowest median life expectancy and Australia and Europe has the highest life expentancy. 


```{r}

# Boxplot life expectancy vs population in different continents

ggplot(data = hw01gapminder) + 
  geom_boxplot(mapping = aes(x = pop, y = lifeExp, color = continent))  + coord_flip()
```

Australia has the smallest range of life expectancy from 1952 - 2007 and the smallest continent population, Asia has the largest range of life Expectancy. It seems the population size does corelate with life expectancies, except in Africa. One possible reason could be due to lower GDP per Capita, therefore has a poorer nationwide health infrastucture.
